\\Large Rising temperatures alter how and where boreal mammals move

### 2025-10-15 writing/manuscript.Rmd
specified "boreal mammals" in title and text

---
title: "\\Large Rising temperatures alter how and where mammals move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Christen H. Fleming
    institute: [ucf, smith]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
# csl: 'movement-ecology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=footnotesize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023. We show that temperature significantly affected how and where these mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century. Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less. Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in phenology. We also predict that rising temperatures will reduce median habitat selection strength for four of the species, although cougars and elk are expected to show increased selection strength for higher altitudes. As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions. Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities. Anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

Across all species, Relative Selection Strength (RSS) was weakest for forest cover and strongest for elevation. At temperatures near 0$^\circ$C, boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km, while southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km. Cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km. Elk selected for intermediate forest cover ($\approx$ 50%), elevations between 1 and 2 km, and distances from water of 10-15 km. Grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km. Mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km. Finally, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km.There was relatively strong agreement between models with and without temperature (Figs. B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B). Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom due to simpler smooth terms) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

\noindent We have demonstrated that temperature is an important driver of how and where large mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways. However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into windows (e.g., day/night and seasons). We were thus able to present results using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed nuanced, continuous-time estimates of change. This framework was essential for estimating how mammals behave at extreme temperatures. In fact, despite the recent increase in frequency of extreme heat events in western Canada, including a severe heat wave in 2021 [@zhang_increased_2023; @thompson_2021_2022; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by HGAMs [@wood_generalized_2017]. Additionally, the hierarchical approach allowed us to leverage data from multiple individuals in a single model while accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully. While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1].

@@

We have demonstrated that temperature has complex and nonlinear effects on how and where mammals move. Rising temperatures have impacted mammals' daily and seasonal movement cycles, and changes are expected to intensify throughout the 21^st^ century. Quantifying how temperature affects mammalian behavior and fitness is a first step towards developing an understanding of the effects of climate change on landscape dynamics and the effectiveness of protected areas in the future. By leveraging the flexibility and statistical power of Hierarchical Generalized Additive Models, we have provided researchers with malleable and data-driven methods for assessing species' responses to climate change. We hope that future work will develop more species-specific models with the intent of developing a community-level framework for quantifying and predicting how communities will respond to climate change.

>>>>

---
title: "\\Large Rising temperatures alter how and where boreal mammals move"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Christen H. Fleming
    institute: [ucf, smith]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: TJ Gooliaff
    institute: [forest]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - forest: British Columbia Ministry of Forests, Lands, Natural Resource Operations and Rural Development, Penticton, BC, Canada.
  - bcp: British Columbia Ministry of Environment and Parks, Penticton, BC, Canada.
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'global-change-biology.csl'
# csl: 'movement-ecology.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=footnotesize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage[left]{lineno} # for line numbers
    - \usepackage{booktabs} # to prevent tables from splitting
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
editor_options: 
  chunk_output_type: console
---

@@

```{r abstract, echo=FALSE}
abstract <- "Widespread warming during the last century has caused many terrestrial mammals to change how and where they move, with cascading effects on fitness and community dynamics. Previous studies have estimated the effects of temperature on mammalian movement behaviour, but few disentangled them from seasonal behaviour cycles. Consequently, it is still uncertain how mammals will adapt their movement behaviour throughout the 21^st^ century. We address this gap by quantifying the effects of temperature on the movement rates and habitat selection of six large-bodied mammalian species (boreal and southern mountain caribou, cougars, elk, grizzly bears, mountain goats, and wolves) throughout western Canada between 1998 and 2023. We show that temperature significantly affected how and where these boreal mammals moved, and we predict that climate change will impact mammals' movement behaviour throughout the 21^st^ century. Projected responses to climate change suggested that rising temperatures will drive southern mountain caribou and mountain goats to move more, while cougars, elk, and wolves will move less. Boreal caribou and grizzly bears showed little change in projected yearly movement rates but clear shifts in phenology. We also predict that rising temperatures will reduce median habitat selection strength for four of the species, although cougars and elk are expected to show increased selection strength for higher altitudes. As mammals increasingly alter their movement rates and select against portions of their current ranges, changes in individuals' movement behaviour will impact encounter rates, including predator-prey dynamics and human-wildlife interactions. Conservation efforts should therefore account for future changes in movement behaviour as well as the consequences such changes may have on communities. Anticipating changes in mammalian movement behaviour will become crucial for effectively and proactively understanding community-level responses and selecting high-quality habitat for long-term conservation."
if(FALSE){
  stringi::stri_count_words(abstract)
}
```

@@

Across all species, Relative Selection Strength (RSS) was weakest for forest cover and strongest for elevation. At temperatures near 0$^\circ$C, boreal caribou selected for forest cover between 50% and 75%, elevations near 500 m, and distances from water < 10 km, while southern mountain caribou selected for dense forest cover, elevations near 2 km, and distances from water $\lessapprox$ 5 km. Cougars selected for dense forest cover (> 75%), an elevation of ~ 1 km, and distances from water < 7.5 km. Elk selected for intermediate forest cover ($\approx$ 50%), elevations between 1 and 2 km, and distances from water of 10-15 km. Grizzly bears selected for relatively sparse forest cover (25-50%), elevation between 1 and 2 km, and distances from water < 3 km. Mountain goats selected for sparse forest cover (< 25%), elevations near 1.5 km, and distances from water < 5 km. Finally, wolves selected for forest cover ($\gtrapprox$ 50%), elevations near 1 km, and distances from water < 5 km.There was relatively strong agreement between models with and without temperature (Figs. B3, and B14), but including temperature always resulted in better fits (all p-values $< 2.2\times10^{-16}$; all $\Delta$AIC $\ge$ 342; Appendix B). Accounting for temperature even resulted in a simpler HRSF for mountain goats (lower model degrees of freedom due to simpler random smooth terms) that fit the data better (lower deviance), and hence no test was necessary to show that the inclusion of temperature produced a better fit.

@@

\noindent We have demonstrated that temperature is an important driver of how and where large boreal mammals move, and that it affects mammals' seasonal and daily movement behaviour in complex and nonlinear ways. However, predicting mammals' responses to climate change remains a challenge, as habitats are warming rapidly and mammals are exposed to increasingly novel and extreme conditions [@ipcc_climate_2023; @diffenbaugh_changes_2013; @botero_evolutionary_2015]. We leveraged the flexibility and interpretability of HGAMs to estimate mammals' nonlinear responses to temperature without imposing rigid assumptions about the shape of the responses [@simpson_method_2025; @simpson_modelling_2018]. In particular, tensor product interaction terms were crucial for estimating smooth, nonlinear changes in daily and seasonal trends with temperature with data from multiple individuals. By allowing interaction terms to be smooth bivariate functions, we were able to quantify changes in movement behaviour without the need to discretize time into windows (e.g., day/night and seasons). We were thus able to present results using relatively simple and interpretable surface plots (Figs. \@ref(fig:dist) and \@ref(fig:hrsfs)) that conveyed nuanced, continuous-time estimates of change. This framework was essential for estimating how mammals behave at extreme temperatures. In fact, despite the recent increase in frequency of extreme heat events in western Canada, including a severe heat wave in 2021 [@zhang_increased_2023; @thompson_2021_2022; @kienzle_has_2018], such events remain relatively rare, so estimating mammals' nonlinear responses required the flexibility provided by HGAMs [@wood_generalized_2017]. Additionally, the hierarchical approach allowed us to leverage data from multiple individuals in a single model while accounting for differences in behaviour and data availability across individuals [@pedersen_hierarchical_2019; @muff_accounting_2020; @mccabe_resource_2021; @jeltsch_need_2025; @silva_too_2025]. Still, GAMs' flexibility and complexity can be a frustrating limitation if used too generously, so we suggest choosing each smooth term's basis size (`k`) carefully. While penalized splines and REML can help avoid unruly smooth terms, the penalty terms can struggle to restrict model complexity when datasets are as large as those presented here [@wood_generalized_2017; @wood_generalized_2017-1].

@@

We have demonstrated that temperature has complex and nonlinear effects on how and where boreal mammals move. Rising temperatures have impacted mammals' daily and seasonal movement cycles, and changes are expected to intensify throughout the 21^st^ century. Quantifying how temperature affects mammalian behavior and fitness is a first step towards developing an understanding of the effects of climate change on landscape dynamics and the effectiveness of protected areas in the future. By leveraging the flexibility and statistical power of Hierarchical Generalized Additive Models, we have provided researchers with malleable and data-driven methods for assessing species' responses to climate change. We hope that future work will develop more species-specific models with the intent of developing a community-level framework for quantifying and predicting how communities will respond to climate change.
